# Aside: OLS via Algebra and Matrices {- .aside-chapter}

## Introduction

Ordinary Least Squares (OLS) regression is a method for estimating the parameters in a linear regression model. This document explains OLS regression first using algebra and then using matrix notation, with a practical implementation in R.

## Algebraic Approach to OLS

In simple linear regression, we model the relationship between a predictor variable $X$ and a response variable $y$ using the equation:

$$ y_i = \beta_0 + \beta_1 X_i + \epsilon_i $$

where:
- $y_i$ is the response variable for observation $i$,
- $X_i$ is the predictor variable for observation $i$,
- $\beta_0$ is the intercept,
- $\beta_1$ is the slope,
- $\epsilon_i$ is the error term.

### Goal: Minimize the Sum of Squared Errors

The OLS method aims to find the values of $\beta_0$ and $\beta_1$ that minimize the sum of squared errors (SSE):

$$SSE = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2$$

where $\hat{y}_i = \beta_0 + \beta_1 X_i$ is the predicted value of $y_i$.

### Deriving the Coefficients

To minimize the SSE, we take the partial derivatives of the SSE with respect to $\beta_0$ and $\beta_1$ and set them to zero:

$$\frac{\partial SSE}{\partial \beta_0} = -2 \sum_{i=1}^{n} (y_i - \beta_0 - \beta_1 X_i) = 0$$
$$\frac{\partial SSE}{\partial \beta_1} = -2 \sum_{i=1}^{n} (y_i - \beta_0 - \beta_1 X_i) X_i = 0$$

Solving these equations gives the OLS estimates:

$$\hat{\beta_1} = \frac{\sum_{i=1}^{n} (X_i - \bar{X})(y_i - \bar{y})}{\sum_{i=1}^{n} (X_i - \bar{X})^2}$$
$$\hat{\beta_0} = \bar{y} - \hat{\beta_1} \bar{X}$$

where $\bar{X}$ and $\bar{y}$ are the means of $X$ and $y$, respectively.

## Matrix Approach to OLS

The algebraic method works well for simple linear regression, but for multiple regression, the matrix approach is more convenient. In matrix notation, the linear model is:

$$\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}$$

where:

* $\mathbf{y}$ is an $n \times 1$ vector of the response variable,
* $\mathbf{X}$ is an $n \times p$ matrix of predictors (with a column of ones for the intercept),
* $\boldsymbol{\beta}$ is a $p \times 1$ vector of coefficients,
* $\boldsymbol{\epsilon}$ is an $n \times 1$ vector of errors.

### Goal: Minimize the Sum of Squared Errors

The OLS method minimizes the sum of squared errors, which in matrix form is:

$$SSE = (\mathbf{y} - \mathbf{X} \boldsymbol{\beta})^T (\mathbf{y} - \mathbf{X} \boldsymbol{\beta})$$

### Deriving the Coefficients

To find the coefficients that minimize the SSE, we set the derivative with respect to $\boldsymbol{\beta}$ to zero:

$$\frac{\partial SSE}{\partial \boldsymbol{\beta}} = -2 \mathbf{X}^T (\mathbf{y} - \mathbf{X} \boldsymbol{\beta}) = 0$$

Solving for $\boldsymbol{\beta}$ gives the OLS estimate:

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \mathbf{X})^{-1} \mathbf{X}^T \mathbf{y}$$

## Practical Implementation in R

### Step 1: Prepare the Data

We'll use a small dataset with one predictor (X) and one response variable (y).

```{r}
# Create a simple dataset
set.seed(123)  # For reproducibility
X <- matrix(rnorm(10), ncol=1)  # 10 random predictors
y <- 3 + 2 * X + rnorm(10)  # Response variable with some noise

# Display the data
data.frame(X = X, y = y)
```

### Step 2: Add an Intercept Column to the Predictor Matrix

We need to add a column of ones to the predictor matrix to account for the intercept term in the regression model.

```{r}
# Add intercept column to the predictor matrix
X_with_intercept <- cbind(1, X)
# Display the new predictor matrix
X_with_intercept
```

### Step 3: Compute the OLS Coefficients

The OLS coefficients can be computed using the formula $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \mathbf{X})^{-1} \mathbf{X}^T \mathbf{y}$.

```{r}
# Compute OLS coefficients
beta <- solve(t(X_with_intercept) %*% X_with_intercept) %*% t(X_with_intercept) %*% y

# Print the coefficients
beta
```

### Step 4: Make Predictions

We can use the computed coefficients to make predictions for the response variable.

```{r}
# Make predictions
y_pred <- X_with_intercept %*% beta

# Print predictions
y_pred
```

### Step 5: Verify with `lm` Function (Optional)

To verify our matrix-based implementation, we can compare the results with the `lm` function in R.

```{r}
# Verify with lm function
model <- lm(y ~ X)
summary(model)
```

## Conclusion

In this document, we explained OLS regression using both algebra and matrix notation. We demonstrated how to derive the OLS estimates and implemented the matrix-based method in R. This approach helps build a strong understanding of the underlying mechanics of linear regression.
